---
title: "Confidence intervals and tests in emmeans"
author: "Russ Lenth"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Confidence intervals and tests}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, echo = FALSE, results = "hide", message = FALSE}
require("emmeans")
knitr::opts_chunk$set(collapse = TRUE, fig.width = 4.5)
```

## Contents {#contents}
This vignette describes various ways of summarizing `emmGrid` objects.

  1. [`emmGrid` objects, described](#emmobj)
  2. [`summary(), `confint()`, and `test()`](#summary)
  3. [Back-transforming to response scale](#tran)
     (See also the ["transformations" vignette](transformations.html))
  4. [Multiplicity adjustments](#adjust)
  5. [Using "by" variables](#byvars)
  6. [Joint (omnibus) tests](#joint)
  7. [Testing equivalence, noninferiority, nonsuperiority](#equiv)
  8. Graphics (in ["basics" vignette](basics.html#plots))
  
[Vignette index](index.html)

## Objects, structures, and methods {#emmobj}
The `ref_grid()` and `emmeans()` functions are introduced in 
[the "Basics" vignette](basics.html). These functions, and a few related ones,
return an object of class `emmGrid`:
```{r}
pigs.lm <- lm(log(conc) ~ source + factor(percent), data = pigs)
pigs.rg <- ref_grid(pigs.lm)
class(pigs.rg)

pigs.emmGrid.s <- emmeans(pigs.rg, "source")
class(pigs.emmGrid.s)
```
If you simply show these objects, you get different-looking results:
```{r}
pigs.rg

pigs.emmGrid.s
```
This is based on guessing what users most need to see when displaying the object.
You can override these defaults; for example,
```{r}
str(pigs.emmGrid.s)
```
The most important method for `emmGrid` objects is `summary()`. It is used as the
default for displaying an `emmeans()` result like `pigs.emmGrid.s`. This `summary()`
method for `emmGrid` objects) actually produces a `data.frame`, but with extra bells and
whistles:
```{r}
class(summary(pigs.emmGrid.s))
```
This can be useful to know because if you want to actually *use* `emmeans()` results
in other computations, you should save its summary, and then you can access those
results just like you would access data in a data frame. The `emmGrid` object itself
is not so accessible. There is a `print.summary_emm()` function that is what
actually produces the output you see above -- a data frame with extra
annotations.

[Back to Contents](#contents)

## `summary()` and its relatives {#summary}
As you may have gathered, the most important method for `emmGrid` objects is `summary()`.
It has a lot of options, and the detailed documentation via 
[`help("summary.emmGrid")`](../html/summary.emmGrid.html) is worth a look.

Just `summary(<object>)` by itself will produce a summary that varies somewhat
according to context. It does this by setting different defaults for the `infer`
argument, which consists of two logical values, specifying confidence intervals 
and tests, respectively. 
The summary of a newly made reference grid will show just estimates and
standard errors, but not confidence intervals or tests (that is, `infer =
c(FALSE, FALSE)`). The summary of an `emmeans()` result, as we see above, will
have intervals, but no tests (i.e., `infer = c(TRUE, FALSE)`); and the result of
a `contrast()` call (see [comparisons and contrasts](comparisons.html)) will
show test statistics and *P* values, but not intervals (i.e., `infer = c(FALSE,
TRUE)`). There are courtesy methods `confint()` and `test()` that just call
`summary()` with the appropriate `infer` setting; for example,
```{r}
test(pigs.emmGrid.s)
```
It is not particularly useful, though, to test these EMMs against the default of 
zero -- which is why tests are not usually shown. It makes a lot more sense
to test them against some target concentration, say 40. And suppose we want to do a one-sided test to see if the concentration is greater than 40. Remembering that the
response is log-transformed in this model, 
```{r}
test(pigs.emmGrid.s, null = log(40), side = ">")
```
[Back to Contents](#contents)


## Back-transforming {#tran}
Transformations and link functions are supported an several ways in **emmeans**,
making this a complex topic worthy of [its own vignette](transformations.html).
Here, we show just the most basic approach. Namely, specifying the argument
`type = "response"` will cause the displayed results to be back-transformed
to the response scale, when a transformation or link function is incorporated
in the model. For example, let's try the preceding `test()` call again:
```{r}
test(pigs.emmGrid.s, null = log(40), side = ">", type = "response")
```
Note what changes and what doesn't change. In the `test()` call, we *still* use
the log of 40 as the null value; `null` must always be specified on the
linear-prediction scale, in this case the log. In the output, the displayed
estimates, as well as the `null` value, are shown back-transformed. As well, the
standard errors are altered (using the delta method). However, the *t* ratios
and *P* values are identical to the preceding results. That is, the tests
themselves are still conducted on the linear-predictor scale (as is noted in the
output).

Similar statements apply to confidence intervals on the response scale:
```{r}
confint(pigs.emmGrid.s, side = ">", level = .90, type = "response")
```
With `side = ">"`, a *lower* confidence limit is computed on the log scale, then
that limit is back-transformed to the response scale. (We have also illustrated 
how to change the confidence level.)

[Back to Contents](#contents)


## Multiplicity adjustments {#adjust}
Both tests and confidence intervals may be adjusted for simultaneous inference. 
Such adjustments ensure that the confidence coefficient for a whole set of 
intervals is at least the specified level, or to control the overall
significance level for a whole family of tests. This is done via the `adjust`
argument. For `ref_grid()` and `emmeans()` results, the default is `adjust =
"none"`. For most `contrast()` results, `adjust` is often something else,
depending on what type of contrasts are created. For example, pairwise
comparisons default to `adjust = "tukey"`, i.e., the Tukey HSD method.
The `summary()` function sometimes *changes* `adjust` if it is inappropriate.
For example, with
```{r}
confint(pigs.emmGrid.s, adjust = "tukey")
```
the adjustment is changed to the Sidak method because the Tukey adjustment is inappropriate unless you are doing pairwise comparisons.

The adjustment method that is *never* inappropriate is Bonferroni; however, it
can be quite conservative. Using `adjust = "mvt"` is the closest to being the
"exact" all-around method, as it uses the multivariate *t* distribution (and the
**mvtnorm** package) with the same covariance structure as the estimates to
determine the adjustment. However, this comes at high computational expense as
the computations are done using simulation techniques. For a large set of tests
(and especially confidence intervals), the computational lag becomes noticeable
if not intolerable.

For tests, `adjust` increases the *P* values over those otherwise obtained with
`adjust = "none"`, making it harder to declare an individual test as
"significant." Compare the following adjusted tests with the unadjusted ones
previously computed.
```{r}
test(pigs.emmGrid.s, null = log(40), side = ">", adjust = "bonferroni")
```

[Back to Contents](#contents)


## "By" variables {#byvars}
Sometimes you want to break a summary down into smaller pieces; for this
purpose, the `by` argument in `summary()` is useful. For example,
```{r}
confint(pigs.rg, by = "source")
```
If there is also an `adjust` in force when `by` variables are used, the 
adjustment is made *separately* on each `by` group; e.g., in the above, we would
be adjusting for sets of 4 intervals, not all 12 together.

There can be a `by` specification in `emmeans()` (or equivalently, a `|` in the
formula); and if so, it is passed on to `summary()` and used unless overridden
by another `by`. Here are examples, not run:
```{r eval = FALSE}
emmeans(pigs.lm, ~ percent | source)     ### same results as above
summary(.Last.value, by = percent)       ### grouped the other way
```
Specifying `by = NULL` will remove all grouping.


[Back to Contents](#contents)


## Joint tests {#joint}
From the above, we already know how to test individual results. For pairwise comparisons (details in [the "comparisons" vignette](comparisons.html)), we might do
```{r}
pigs.prs.s <- pairs(pigs.emmGrid.s)
pigs.prs.s
```
But suppose we want an *omnibus* test that all these comparisons are zero.
Easy enough, using the `joint` argument in `test` (note: *not* available 
in `summary()`):
```{r}
test(pigs.prs.s, joint = TRUE)
```
Notice that there are three comparisons, but only 2 d.f. for the test,
as cautioned in the message.

[Back to Contents](#contents)


## Testing equivalence, noninferiority, and nonsuperiority {#equiv}
The `delta` argument in `summary()` or `test()` allows the user to
specify a threshold value to use in a test of equivalence, noninferiority, or nonsuperiority. An equivalence test is kind of a backwards significance test, where differences enough smaller than `delta` are the ones that can be significant.  The help page for `summary.emmGrid` gives the details of these tests.
Suppose in the present example, we consider two sources to be equivalent if they 
are within 25% of each other. We can test this as follows:
```{r}
test(pigs.prs.s, delta = log(1.25), adjust = "none")
```
By our 25% standard, soy and skim are equivalent at the $\alpha = .05$ level,
when no multiplicity adjustment is used.

[Back to Contents](#contents)


## Graphics {#graphics}
Graphical displays of `emmGrid` objects are described in the 
["basics" vignette](basics.html#plots)

