---
title: "Sophisticated models in emmeans"
author: "Russ Lenth"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sophisticated models in emmeans}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, echo = FALSE, results = "hide", message = FALSE} 
require("emmeans") 
options(show.signif.stars = FALSE) 
knitr::opts_chunk$set(collapse = TRUE,
fig.width = 4.5) 
```

This vignette gives a few examples of the use of the **emmeans** package to
analyze other than the basic types of models provided by the **stats**
package. Emphasis here is placed on accessing the optional capabilities
that are typically not needed for the more basic models. A reference for
all supported models is provided in the ["models" vignette](models.html).

## Contents {#contents}

 1. [Linear mixed models (lmer)](#lmer)
    a. [System options for lmerMod models](#lmerOpts)
 2. [Ordinal models](#ordinal)
 3. [Models fitted using MCMC methods](#mcmc)
 4. [Reference for supported models](models.html)
 
[Vignette index](index.html) 


## Linear mixed models (lmer) {#lmer}
Linear mixed models are really important in statistics. Emphasis here is placed
on those fitted using `lme4::lmer()`, but **emmeans** also supports other
mixed-model packages such as **nlme**.

To illustrate, consider the `Oats` dataset in the **nlme** package. It has the
results of a balanced split-plot experiment: experimental blocks are divided 
into plots that are randomly assigned to oat varieties, and the plots are
subdivided into subplots that are randomly assigned to amounts of nitrogen
within each plot. We will consider a linear mixed model for these data, 
excluding interaction (which is justified in this case). For sake of 
illustration, we will exclude a few observations.
```{r}
Oats.lmer <- lme4::lmer(yield ~ Variety + factor(nitro) + (1|Block/Variety),
                        data = nlme::Oats, subset = -c(1,2,3,5,8,13,21,34,55))
```
Let's look at the EMMs for `nitro`:
```{r}
Oats.emm.n <- emmeans(Oats.lmer, "nitro")
Oats.emm.n
```
You will notice that the degrees of freedom are fractional: that is due to the
fact that whole-plot and subplot variations are combined when standard errors
are estimated. Different degrees-of-freedom methods are available. By default, 
the Kenward-Roger method is used, and that's why you see a message about the
**pbkrtest** package being loaded, as it implements that method. We may specify
a different degrees-of-freedom method via the optional argument `lmer.df`:
```{r}
emmeans(Oats.lmer, "nitro", lmer.df = "satterthwaite")
```
This latest result uses the Satterthwaite method, which is implemented in the
**lmerTest** package. Note that, with this method, not only are the degrees of
freedom slightly different, but so are the standard errors. That is because the 
Kenward-Roger method also entails making a bias adjustment to the covariance
matrix of the fixed effects; that is the principal difference between the
methods. A third possibility is `"asymptotic"`:
```{r}
emmeans(Oats.lmer, "nitro", lmer.df = "asymptotic")
```
This just sets all the degrees of freedom to `NA` -- that's **emmeans**'s way of
using *z* statistics rather than *t* statistics. The asymptotic methods tend to
make confidence intervals a bit too narrow and P values a bit too low; but they
involve much, much less computation. Note that the SEs are the same as obtained
using the Satterthwaite method.

Comparisons and contrasts are pretty much the same as with other models. As
`nitro` has quantitative levels, we might want to test polynomial contrasts:
```{r}
contrast(Oats.emm.n, "poly")
```
The interesting thing here is that the degrees of freedom are much larger than
they are for the EMMs. The reason is because `nitro` within-plot factor, so
inter-plot variations have little role in estimating contrasts among `nitro`
levels. On the other hand, `Variety` is a whole-plot factor, and there is not 
much of a bump in degrees of freedom for comparisons:
```{r}
emmeans(Oats.lmer, pairwise ~ Variety)
```

### System options for lmerMod models {#lmerOpts}
The computation required to compute the adjusted covariance matrix and degrees 
of freedom may become cumbersome. Some user options (i.e., `emm_options()` 
calls) make it possible to streamline these computations through default methods
and limitations on them. First, the option `lmer.df`, which may have values of 
`"kenward-roger"`, `"satterthwaite"`, or `"asymptotic"` (partial matches are 
OK!) specifies the default degrees-of-freedom method.

The options `disable.pbkrtest` and `disable.lmerTest` may be `TRUE` or `FALSE`,
and comprise another way of controlling which method is used (e.g., the
Kenward-Roger method will not be used if `get_emm_option("disable.pbkrtest") ==
TRUE`). Finally, the options `pbkrtest.limit` and `lmerTest.limit`, which should
be set to numeric values, enable the given package conditionally on whether the
number of data rows does not exceed the given limit. The factory default is 3000
for both limits.

[Back to Contents](#contents)


## Ordinal models {#ordinal}
Ordinal-response models comprise an example where several options are 
available for obtaining EMMs. To illustrate, consider the `wine` data
in the **ordinal** package. The response is a rating of bitterness on
a five-point scale. we will consider a probit model in two factors
during fermentation: `temp` (temperature) and `contact` (contact with 
grape skins), with the judge making the rating as a scale predictor:
```{r}
require("ordinal")
wine.clm <- clm(rating ~ temp + contact, scale = ~ judge,
                data = wine, link = "probit")
```
(in earlier modeling, we found little interaction between the factors.)
Here are the EMMs for each factor using default options:
```{r}
emmeans(wine.clm, list(pairwise ~ temp, pairwise ~ contact))
```
These results are on the "latent" scale; the idea is that there is a continuous 
random variable (in this case normal, due to the probit link) having a mean that
depends on the predictors; and that the ratings are a discretization of the 
latent variable based on a fixed set of cut points (which are estimated). In 
this particular example, we also have a scale model that says that the variance 
of the latent variable depends on the judges. The latent results are quite a bit
like those for measurement data, making them easy to interpret. The only catch 
is that they are not uniquely defined: we could apply a linear transformation to
them, and the same linear transformation to the cut points, and the results
would be the same.

The `clm` function actually fits the model using an ordinary probit model
but with different intercepts for each cut point. We can get detailed 
information for this model by specifying `mode = "linear.predictor"`:
```{r}
tmp <- ref_grid(wine.clm, mode = "lin")
tmp
```
Note that this reference grid involves an additional constructed predictor named
`cut` that accounts for the different intercepts in the model. Let's obtain EMMs
for `temp` on the linear-predictor scale:
```{r}
emmeans(tmp, "temp")
```
These are just the negatives of the latent results obtained earlier (the sign is
changed to make the comparisons go the right direction). Closely related to this
is `mode = "cum.prob"` and `mode = "exc.prob"`, which simply transform the 
linear predictor to cumulative probabilities and exceedance (1 - cumulative) 
probabilities. These modes give us access to the details of the fitted model but
are cumbersome to use for describing results. When they can become useful is
when you want to work in terms of a particular cut point. Let's look at `temp`
again in terms of the probability that the rating will be at least 4:
```{r}
emmeans(wine.clm, ~ temp, mode = "exc.prob", at = list(cut = "3|4"))
```

There are yet more modes! With `mode = "prob"`, we obtain estimates of the probability distribution of each rating. Its reference grid includes a factor
with the same name as the model response -- in this case `rating`. We usually 
want to use that as the primary factor, and the factors of interest as `by` variables:
```{r}
emmeans(wine.clm, ~ rating | temp, mode = "prob")
```
Using `mode = "mean.class"` obtains the average of these probability 
distributions as probabilities of the integers 1--5:
```{r}
emmeans(wine.clm, "temp", mode = "mean.class")
```

And there is a mode for the scale model too. In this example, the scale model
involves only judges, and that is the only factor in the grid:
```{r}
summary(ref_grid(wine.clm, mode = "scale"), type = "response")
```
Judge 8's ratings don't vary much, relative to the others.
The scale model is in terms of log(SD). Again, these are not uniquely
identifiable, and the first level's estimate is set to log(1) = 0. so,
actually, each estimate shown is a comparison with judge 1.

[Back to Contents](#contents)


## Models fitted using MCMC methods {#mcmc}
To illustrate **emmeans**'s support for models fitted using MCMC methods,
consider the `example_model` available in the **rstanarm** package. 
The example concerns CBPP, a serious disease of cattle in Ethiopia.
A generalized linear mixed model was fitted to the data using the
code below. We subsequently obtain its reference grid in the usual way.
<!--- I'm faking this; I actually saved the ref_grid in a system file --->
```{r eval = FALSE}
example_model <- rstanarm::stan_glmer(
    cbind(incidence, size - incidence) ~ size + period + (1|herd),
    data = lme4::cbpp, family = binomial,
    chains = 2, cores = 1, seed = 12345, iter = 500)
rst_ex.rg <- ref_grid(example_model)
```
<!--- here's the system file with the ref_grid --->
```{r echo = FALSE}
load(system.file("extdata", "rstex.RData", package = "emmeans"))
rst_ex.rg <- do.call(emmobj, rstex)
```
Here is the structure of the reference grid:
```{r}
rst_ex.rg
```
And, again in the usual way, we can obtain EMMs:
```{r}
period.emm <- emmeans(rst_ex.rg, "period")
period.emm
```
The summary shows a frequentist summary of the EMMs. Under the hood, though, is a posterior sample of 500 sets of EMMs -- each one being the EMMs computed from each set of parameter values in the posterior sample of regression coefficients. The summary we see is based on the sample means and covariances of that sample.

We can access these posterior results via the `as.mcmc` method for `emm` objects.
This gives us an object of class `mcmc` (defined in the **coda** package), which
can be summarized and explored as we please.
```{r}
require("coda")  ### needed to access generic for as.mcmc()
summary(as.mcmc(period.emm))
```
Note that `as.mcmc` will actually produce an `mcmc.list` when there is more than
one chain present, as in this example.
The 2.5th and 97.5th quantiles are similar, but not identical, to the
95% confidence intervals in the frequentist summary. Here is a plot of
the posterior EMMs, back-transformed:
```{r}
bayesplot::mcmc_areas(as.mcmc(regrid(period.emm)))
```

... and here are intervals for each period compared with its neighbor; we wrap 
the important part of the call in an extra `as.mcmc(as.matrix(...))` so as to
combine the two chains into one.
```{r}
HPDinterval(as.mcmc(as.matrix(
    as.mcmc(contrast(period.emm, "consec", reverse = TRUE), names = FALSE))))
```
The only interval that excludes zero is the one that compares periods 1 and 2.

In summary, to do Bayesian analysis in the **emmeans** package, use the same tools that are available for other models, extract the MCMC samples using `as.mcmc()`, and summarize or plot from there.

[Back to Contents](#contents)
